Experiences with Planning for Natural Language Generation

### 2009-06-02 trunk/Papers/ci-spark/domains.tex
changes to section 3

We will now focus the discussion on two specific NLG problems:
sentence generation in the sense of \citet{KolSto07}, and the
generation of instructions in virtual environments
\citep{ByrKolStrCasDalMooObe09}. For each, we will introduce the task
and show by example how the task can be seen as a planning problem.

@@

An example is the tree-adjoining grammar (TAG;
\citealt{joshi;etal1997}) shown in
Figure~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists of
\emph{elementary trees} (i.e., the disjoint trees in the figure), each
of which contributes certain \emph{semantic content}. Let's say that
we assume a knowledge base containing the individuals $e$, $r_1$ and
$r_2$ and the facts that $r_1$ and $r_2$ are rabbits, $r_1$ is white
and $r_2$ is brown, and $e$ is an event in which $r_1$ sleeps. Then we
would want to construct a sentence expressing the information
$\{\mathsf{sleep}(e,r_1)\}$ by combining instances of the elementary
trees (in which the \emph{semantic roles}, such as $\mathsf{self}$ and
$\mathsf{subj}$, have been substituted by constants from the knowledge
base) into a TAG derivation as as shown in
Figure~\ref{fig:white-rabbit-sleeps-deriv}. In the picture, the dashed
arrow indicates TAG's \emph{substitution} operation, which ``plugs''
an elementary into the leaf of another tree, and the dotted arrows
stand for \emph{adjunction}, which splices an elementary tree into an
internal node. We can then read off the sentence ``The white rabbit
sleeps'' from the derivation. Note that the sentence ``The rabbit
sleeps'' would not have been an appropriate result, because ``the
rabbit'' could refer to either $r_1$ or $r_2$, and thus $r_2$ remains
as a \emph{distractor}, i.e.\ an incorrect possible interpretation of
the phrase.

@@

This perspective on sentence generation has the advantage of solving
the sentence planning and surface realization problems
simultaneously. The generation of referring expressions (REs) is
usually seen as a sentence planning task; in the example, we required
that the referring expression ``the white rabbit'' can be resolved
uniquely to $r_1$ by the hearer, in addition to the requirement that
the derivation is grammatically correct. This is useful in cases where
sentence planning and surface realization interact, because syntactic
information about the individual words is available when the REs are
generated (see e.g. \citealt{stone98textual}). However, the problem of
whether a given communicative goal can be achieved with a given
grammar is NP-complete \citep{KolStr02}. A naive search algorithm that
computes the derivation top-down takes exponential time and is clearly
infeasible to use in practice. The seminal SPUD system
\citep{Stone2003a}, which first worked out the idea of integrated
TAG-based sentence generation, used a greedy algorithm to cope with
the search, which circumvents the combinatorial explosion but is
incomplete.

@@

As an alternative, \citet{KolSto07} recently proposed to control the
search by converting the sentence generation problem into a planning
problem and then running a planner \citet{KolSto07}.\footnote{See
  \url{http://code.google.com/p/crisp-nlg/} for the CRISP system, in
  which this conversion is implemented.} This planning problem assumes
an initial state containing an atom $\mathsf{subst}(S,\mathsf{root})$,
encoding that we must generate a sentence ($S$) starting at the node
named $\mathsf{root}$ in the TAG derivation tree, and an atom
$\mathsf{referent}(\mathsf{root},e)$ encoding that the (event)
individual which the entire sentence describes is $e$. Then it encodes
the addition of elementary trees to the TAG derivation in the
individual planning operators; the ones that are needed for our
example above are shown in
Fig.~\ref{fig:white-rabbit-as-planning}.

Here the action instance $\addact{sleeps}(\mathsf{root}, e, r_1)$
replaces the atom $\mathsf{subst}(S,\mathsf{root})$ with the atom
$\mathsf{subst}(NP,\mathsf{subj}(\mathsf{root}))$. In an abuse of PDDL
syntax, we write $\mathsf{subj}(\mathsf{root})$ as shorthand for a
fresh individual name.\footnote{These terms, which are not valid in
  ordinary PDDL, can be eliminated by estimating an upper bound $n$
  for the plan length, making $n$ copies of each action, ensuring that
  copy $i$ can only be applied in step $i$, and replacing the term
  $\mathsf{subj}(u)$ in an action copy by the constant
  $\mathsf{subj}_i$. The terms $S$, $NP$, and $N$ in the planning
  problem are constants.}  At the same time, the operator records that
the semantic information $\mathsf{sleep}(e,r_1)$ has now been
expressed, and introduces all individuals except for $r_1$ as
distractors for the new RE at $\mathsf{subj}(\mathsf{root})$. These
distractors can then be removed by subsequent applications of the
other two operators. Eventually we reach a goal state, which is
characterized by goals including $\forall x \forall y. \neg
\mathsf{subst}(x,y)$, $\forall x \forall y. \neg
\mathsf{distractor}(x,y)$, and
$\mathsf{expressed}(\mathsf{sleep},e,r_1)$. In summary, the following
is a correct plan:
%
\begin{enumerate}
\item $\mathsf{add}\textsf{-}\mathsf{sleeps}(\mathsf{root}, r_1)$,
\item $\mathsf{add}\textsf{-}\mathsf{rabbit}(\mathsf{subj}(\mathsf{root}),r_1)$,
\item $\mathsf{add}\textsf{-}\mathsf{white}(\mathsf{subj}(\mathsf{root}),r_1)$.
\end{enumerate}
%
The grammatical derivation in
Figure~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the
generated sentence ``the white rabbit sleeps'', can be systematically
reconstructed from this plan. Thus, we can solve the sentence
generation problem via the detour through planning and bring current
search heuristics for planning to bear on generation.

>>>>

We begin by considering two specific NLG problems: sentence generation in
the sense of \citet{KolSto07}, and the generation of instructions in
virtual environments \citep{ByrKolStrCasDalMooObe09}. For each problem, we
introduce the task and show by example how the task can be viewed as a
planning problem.

@@

An example of such a lexicalized grammar is the tree-adjoining grammar
(TAG; \citealt{joshi;etal1997}) shown in
Figure~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists of
\emph{elementary trees} (i.e., the disjoint trees in the figure), each of
which contributes certain \emph{semantic content}. For instance, say that a
knowledge base contains the individuals $e$, $r_1$ and $r_2$, and the facts
that $r_1$ and $r_2$ are rabbits, $r_1$ is white and $r_2$ is brown, and
$e$ is an event in which $r_1$ sleeps. We could then construct a sentence
expressing the information $\{\mathsf{sleep}(e,r_1)\}$ by combining
instances of the elementary trees (in which the \emph{semantic roles}, such
as $\mathsf{self}$ and $\mathsf{subj}$, have been substituted by constants
from the knowledge base) into a TAG derivation as shown in
Figure~\ref{fig:white-rabbit-sleeps-deriv}. In the figure, the dashed arrow
indicates TAG's \emph{substitution} operation, which ``plugs'' an
elementary into the leaf of another tree; the dotted arrows stands for
\emph{adjunction}, which splices an elementary tree into an internal node.
We can then read the sentence ``The white rabbit sleeps'' from the
derivation. Note that the sentence ``The rabbit sleeps'' would not have
been an appropriate result, because ``the rabbit'' could refer to either
$r_1$ or $r_2$. Thus, $r_2$ remains as a \emph{distractor}, i.e., an
incorrect possible interpretation of the phrase.

@@

This perspective on sentence generation also has the advantage of solving
the sentence planning and surface realization problems simultaneously,
which is particularly useful in cases where these two problems interact.
For instance, the generation of referring expressions (REs) is usually seen
as a sentence planning task, however, syntactic information about
individual words in available when the REs are generated (see, e.g.,
\citealt{stone98textual}). (In the example, we require the referring
expression ``the white rabbit'' to be resolved uniquely to $r_1$ by the
hearer, in addition to the requirement that the derivation be grammatically
correct.)

@@

However, the problem of deciding whether a given communicative goal can be
achieved with a given grammar is NP-complete \citep{KolStr02}: a naive
search algorithm that computes a derivation top-down takes exponential time
and is clearly infeasible to use in practice. In order to circumvent this
combinatorial explosion, the seminal SPUD system \citep{Stone2003a}, which
first established the idea of integrated TAG-based sentence generation,
used a greedy, but incomplete, search algorithm.
%
To better control the search, \citet{KolSto07} recently proposed an
alternative approach which converts the sentence generation problem into a
planning problem, and solves the transformed search problem using a planner
\citep{KolSto07}.\footnote{See
  \url{http://code.google.com/p/crisp-nlg/} for the CRISP system, in
  which this conversion is implemented.} The resulting planning problem in
this case assumes an initial state containing an atom
$\mathsf{subst}(S,\mathsf{root})$, encoding the fact that a sentence ($S$)
must be generated starting at the node named $\mathsf{root}$ in the TAG
derivation tree, and a second atom $\mathsf{referent}(\mathsf{root},e)$
which encodes the fact that the entire sentence describes the (event)
individual $e$. The elementary trees in the TAG derivation are encoded as
individual planning operators.

Figure~\ref{fig:white-rabbit-as-planning} shows the
transformed planning operators needed to generate the above example
sentence, ``The white rabbit sleeps.'' Here the action instance
$\addact{sleeps}(\mathsf{root}, e, r_1)$ replaces the atom
$\mathsf{subst}(S,\mathsf{root})$ with the atom
$\mathsf{subst}(NP,\mathsf{subj}(\mathsf{root}))$. In an abuse of PDDL
syntax, we write $\mathsf{subj}(\mathsf{root})$ as a shorthand for a fresh
individual name.\footnote{These terms are not valid in
  ordinary PDDL but can be eliminated by estimating an upper bound $n$
  for the plan length, making $n$ copies of each action, ensuring that
  copy $i$ can only be applied in step $i$, and replacing the term
  $\mathsf{subj}(u)$ in an action copy by the constant
  $\mathsf{subj}_i$. The terms $S$, $NP$, and $N$ in the planning
  problem are constants.}  At the same time, the operator records that
the semantic information $\mathsf{sleep}(e,r_1)$ has now been
expressed, and introduces all individuals except $r_1$ as
distractors for the new RE at $\mathsf{subj}(\mathsf{root})$. These
distractors can then be removed by subsequent applications of the
other two operators. Eventually we reach a goal state, which is
characterized by goals including $\forall x \forall y. \neg
\mathsf{subst}(x,y)$, $\forall x \forall y. \neg
\mathsf{distractor}(x,y)$, and
$\mathsf{expressed}(\mathsf{sleep},e,r_1)$. For instance, the following
plan correctly performs the necessary derivation:
%
\begin{enumerate}
\item $\mathsf{add}\textsf{-}\mathsf{sleeps}(\mathsf{root}, r_1)$,
\todo{FIX: add-sleeps takes 3 arguments in the PDDL spec!}
\item $\mathsf{add}\textsf{-}\mathsf{rabbit}(\mathsf{subj}(\mathsf{root}),r_1)$,
\item $\mathsf{add}\textsf{-}\mathsf{white}(\mathsf{subj}(\mathsf{root}),r_1)$.
\end{enumerate}
%
The grammatical derivation in
Figure~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the
generated sentence ``the white rabbit sleeps,'' can be systematically
reconstructed from this plan. Thus, we can solve the sentence
generation problem via the detour through planning and bring current
search heuristics for planning to bear on generation.
